---
title: "Dynamic ATAC-peaks in KRasG12D colonic epithelium using merged peaks with Csnk2a1 subset"
author: "Bing Shui"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: yes
    toc_float: true
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Library loading and set up
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(DESeq2)
library(knitr)
library(kableExtra)
library(plyranges)
library(stringr)
library(biobroom)
library(ggrepel)
library(RColorBrewer)
library(gplots)
library(biomaRt)
library(ChIPseeker)
library(TxDb.Mmusculus.UCSC.mm10.knownGene)
library(mixOmics)
library(mosaic)
library(clusterProfiler)
library(DOSE)
library(org.Mm.eg.db)
library(pathview)
library(AnnotationDbi)
library(fgsea)
library(rvest)
library(ggseqlogo)
library(gridExtra)
library(rtracklayer)
library(Biostrings)
library(Rsamtools)
library(data.table)
library(grid)
library(apeglm)
```

# Load dataset
This dataset is generated using WT and KRas-G12D expressingn colonic epithelium from mouse. Mice user are
 - Fabp-Cre/+
 - Fabp-Cre/+; KRasG12D/+

Alignment and peak calling was done using the `bcbio` ATAC-Seq pipeline following instructions described here: (https://bcbio-nextgen.readthedocs.io/en/latest/contents/atac.html)

Merged peak data are used here. This is obtained by combining all peaks from all samples, merging peaks within 500bp of each other, and using `featureCount` to count all reads in the peaks.

```{r message=FALSE, warning=FALSE}
metadata = readr::read_csv("Data/bcbio_output/metadata.csv") %>%
  dplyr::rename("sample"=...1) %>%
  dplyr::select(-batch, -phenotype) %>%
  dplyr::filter(tissue == "colon_epithelium") %>%
  dplyr::filter(sample != "Kras_colon_rep3") %>%
  mutate(condition=relevel(factor(condition), ref="WT")) %>%
  as.data.frame()
rownames(metadata) = metadata$sample
counts_all = readr::read_tsv("Data/bcbio_output/merged/merged-counts_csnk2a1.tsv") %>%
  tibble::column_to_rownames("id")

```

Metadata
```{r metadata-table}
metadata %>%
  kable() %>%
  kable_styling()
```

Load Ensembl ID dataset
```{r eval = F}
ensembl = useMart("ensembl", dataset="mmusculus_gene_ensembl")
symbols = getBM(attributes=c("entrezgene_id", "mgi_symbol", "ensembl_gene_id"), mart=ensembl) %>%
  dplyr::rename("geneId"=entrezgene_id) %>%
  mutate(geneId=as.character(geneId))

save(symbols, file = "Data/symbols.rda")
```


# Cleanup
We need to do a little it of cleanup before contining. We need to remove peaks
that appear in regions known to be false positive machines, called the blacklist
regions. We also are interested in regions that could be affecting expression
via chromatin accessability, so we'll only consider peaks within a window around
the transcription start sites of genes. 

## Blacklist region removal
Turn out the peak at the TSS of Csnk2a1 falls into the blacklist because that region is considered high signal. I need to manually add it back after blacklist filtering.

```{r remove-blacklist, message=FALSE, warning=FALSE}
load("Data/symbols.rda")

expand_region_string = function(df, column) {
  df = as.data.frame(df)
  tokens = str_split_fixed(df[,column], ":", 2)
  chrom = tokens[,1]
  start = as.numeric(str_split_fixed(tokens[,2], "-", 2)[, 1])
  end = as.numeric(str_split_fixed(tokens[,2], "-", 2)[, 2])
  regions = data.frame(seqnames=chrom, start=start, end=end)
  df %>%
    bind_cols(regions)
}
blacklist = readr::read_tsv("Data/mm10-blacklist.v2.bed.gz", col_names=c("seqnames", "start", "end")) %>% 
  as_granges()
blacklist_peaks = counts_all %>%
  tibble::rownames_to_column("peak") %>%
  expand_region_string("peak") %>%
  as_granges() %>%
  join_overlap_inner(blacklist) %>%
  tidy() %>%
  pull(peak)
counts = counts_all[!rownames(counts_all) %in% blacklist_peaks, ]
counts <- rbind(counts, counts_all[291429,])
counts <- counts[,-6]
```

A few peaks overlap the ENCODE blacklist regions, so we removed them from the analysis. There 
were `r length(blacklist_peaks)` that overlapped with the blacklist regions.

## Open region annotation
Here we annotate the peaks that we called with contextual genomic information. We can also see that
some peaks are annotated in multiple regions, this can't be helped as genes overlap so there
are areas where multiple features of genes overlap.

```{r}
annotatedobj = ChIPseeker::annotatePeak(counts %>%
                                        tibble::rownames_to_column("peak") %>%
                                        expand_region_string("peak") %>%
                                        as_granges(), TxDb = TxDb.Mmusculus.UCSC.mm10.knownGene)
plotAnnoPie(annotatedobj)

```


## Remove peaks not near TSS
Generally we only care about peaks that are close to gene, here we are pretty lenient and consider
peaks that are within 1000 bases of a transcription start site of a gene.

```{r tss-peaks}
annotated = annotatedobj %>%
  as.GRanges() %>%
  tidy() 
ggplot(annotated, aes(distanceToTSS)) +
  stat_density(geom="line") +
  xlim(c(-10000,10000)) 
```

```{r keep-tss-peaks}
KEEP_DISTANCE = 1000
close_to_tss = annotated %>%
  dplyr::filter(abs(distanceToTSS) < KEEP_DISTANCE) %>%
  pull(peak)
counts = counts[rownames(counts) %in% close_to_tss,]
```
Most peaks fall close to the TSS, but there are some that are very far away from the TSS. We'll 
remove peaks that aren't anywhere near the TSS for a gene from the analysis. We'll keep only
peaks within `r KEEP_DISTANCE` bases of a TSS. This leaves us with `r length(close_to_tss)` peaks to consider.

## Post filtering peak annotation
Here we annotate the peaks post-filtering

```{r}
annotatedobj = ChIPseeker::annotatePeak(counts %>%
                                        tibble::rownames_to_column("peak") %>%
                                        expand_region_string("peak") %>%
                                        as_granges(), TxDb = TxDb.Mmusculus.UCSC.mm10.knownGene)
plotAnnoPie(annotatedobj)
upsetplot(annotatedobj)

```

# Differential affinity analysis
Here we look at the **condition_Kras_vs_WT** coefficient, which will show us the peak 
affinity differences between the mouse KRasG12D and WT pancreatic epithelial cell samples.
```{r}
dds = DESeqDataSetFromMatrix(counts, metadata, design=~condition)
dds = DESeq(dds)
# Print scale factors for BW normalization
1/sizeFactors(dds)
plotDispEsts(dds)
CUTOFF = 0.05
dds_result <- lfcShrink(dds, coef = "condition_Kras_vs_WT", type = "normal")
```

## QC
```{r}
DESeq2::plotMA(dds_result)

dds_transform <- varianceStabilizingTransformation(dds)
rawCountTable_transform <- as.data.frame(assay(dds_transform))
pseudoCount_transform = log2(rawCountTable_transform + 1)
```

The top 500 most variable genes are selected for PCA analysis.
```{r}
plotPCA(dds_transform, intgroup = "condition", ntop = 500)

```

## Output results
Signal to noise ratio is calculated using the definition given by Broad [GSEA](https://www.gsea-msigdb.org/gsea/doc/GSEAUserGuideFrame.html?Run_GSEA_Page).
```{r forebrain-hindbrain}
markup_deseq2 = function(res) {
  resanno = res %>%
    expand_region_string("peak") %>%
    as_granges()
  ChIPseeker::annotatePeak(resanno, TxDb=TxDb.Mmusculus.UCSC.mm10.knownGene) %>%
    as.GRanges() %>%
    tidy() %>%
    left_join(symbols, by="geneId")
}

res = dds_result %>%
  as_tibble(rownames = "peak")

# calculate signal-to-noise ratio for GSEA later
s2n <- function(num_list, cond_1 = c(4:5), cond_2 = c(1:3)) {
  mean1 <- mean(num_list[cond_1])
  if (mean1 == 0) {
    mean1 = 1
  }
  mean2 <- mean(num_list[cond_2])
  if (mean2 == 0) {
    mean2 = 1
  }
  sd1 <- sd(num_list[cond_1])
  sd2 <- sd(num_list[cond_2])
  sd1 <- min(sd1, 0.2*abs(mean1))
  sd2 <- min(sd2, 0.2*abs(mean2))
  s2nvalue <- (mean1-mean2)/(sd1+sd2)
  return(s2nvalue)
}

rawCountTable <- as.data.frame(DESeq2::counts(dds, normalize = TRUE)) 

res$s2n <- apply(rawCountTable,1,s2n)

res <- markup_deseq2(res)

res <- res[!duplicated(res$peak),]

write.csv(res, "Result/ATAC_ce_Kras-WT_merged_Csnk2a1subset.csv", row.names = FALSE)

rawCountTable <- rawCountTable %>% rownames_to_column(var = "peak")
rawCountTable <- left_join(rawCountTable, res[,c(6,24)], by = c("peak" ="peak")) 

write.csv(rawCountTable, "Result/ATAC_ce_normalized_count_Csnk2a1 subset.csv", row.names = FALSE, na = "")
```
### Description of output files

- **start**: start of peak
- **end**: end of peak
- **width**: width of peak
- **strand**: strand of peak
- **seqname**: chromosome of peak
- **peak**: ID of peak
- **baseMean**: mean count of reads at peak for all samples
- **estimate**: log2 fold change of comparison
- **stderror**: standard error of the log2 fold change
- **statistic**: value of test statistic (Wald test)
- **p.value**: unadjusted p-value
- **p.adjusted**: adjusted p-value by BH correction
- **annotation**: proximity to nearest gene feature
- **geneChr**: chromosome of gene
- **geneStart**: start coordinate of gene
- **geneEnd**: end coordinate of gene
- **geneId**: Entrez gene ID of gene
- **transcriptId**: transcript ID
- **distanceToTSS**: distance to nearest transcription start site
- **mgi_symbol**: gene symbol from MGI


# Session Info
```{r}
sessionInfo()
```